---
title: "Fast Kalman Filter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast Kalman Filter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                      collapse = TRUE,
                      comment = "#>"
)
```

To demonstrate the use of the `FKF` package this example shows how to fit an
ARMA(2, 1) model using this Kalman filter implimentation and teh extraction of
the resulting filtered and smoothed estimates.

Start by loading the package:
```{r, load_package}
library('FKF')
```

# Generating synthetic data

A data set of 1000 points is generated from the model
\[
a_{t} = \frac{-0.2}{1 - 0.6z^{-1} - 0.2z^{-1}}\eta_t
\]
where $\eta_{t} \sim N\left(0,0.2\right)$

```{r ex1_setup}
n <- 1000

## Set the AR parameters
ar1 <- 0.6
ar2 <- 0.2
ma1 <- -0.2
sigma <- sqrt(0.2)

a <- arima.sim(model = list(ar = c(ar1, ar2), ma = ma1), n = n,
               innov = rnorm(n) * sigma)
```

# Estimation

To estimate the parameters using numerical optimisation two functions are
specified. The first creates a state space representation out of the four ARMA parameters

```{r ex1_ss}
arma21ss <- function(ar1, ar2, ma1, sigma) {
    Tt <- matrix(c(ar1, ar2, 1, 0), ncol = 2)
    Zt <- matrix(c(1, 0), ncol = 2)
    ct <- matrix(0)
    dt <- matrix(0, nrow = 2)
    GGt <- matrix(0)
    H <- matrix(c(1, ma1), nrow = 2) * sigma
    HHt <- H %*% t(H)
    a0 <- c(0, 0)
    P0 <- matrix(1e6, nrow = 2, ncol = 2)
    return(list(a0 = a0, P0 = P0, ct = ct, dt = dt, Zt = Zt, Tt = Tt, GGt = GGt,
                HHt = HHt))
}
```

The second is the objective function passed to `optim` which returns the
noegative log-likelihood

```{r ex1_obj}
objective <- function(theta, yt) {
    sp <- arma21ss(theta["ar1"], theta["ar2"], theta["ma1"], theta["sigma"])
    ans <- fkf(a0 = sp$a0, P0 = sp$P0, dt = sp$dt, ct = sp$ct, Tt = sp$Tt,
               Zt = sp$Zt, HHt = sp$HHt, GGt = sp$GGt, yt = yt)
    return(-ans$logLik)
}
```

Estimation is then performed using a numeric search by `optim`

```{r ex1_opt}
theta <- c(ar = c(0, 0), ma1 = 0, sigma = 1)
fit <- optim(theta, objective, yt = rbind(a), hessian = TRUE)
```

# Results

First consider the 95% CI for the parameter estimate which can be constructed
using the hessian of the fit:

```{r ex1_ci}
## Confidence intervals
p <- cbind(actual = c(ar1=ar1,ar2=ar2,ma1=ma1,sigma=sigma),
           estimate = fit$par,
           lowerCI = fit$par - qnorm(0.975) * sqrt(diag(solve(fit$hessian))),
           upperCI = fit$par + qnorm(0.975) * sqrt(diag(solve(fit$hessian))))
p
```

showing the actual parameters fall within the 95% CI of the estimated values.

The series can be filtered usinf the estimated parameter values
```{r filter}
sp <- arma21ss(fit$par["ar1"], fit$par["ar2"], fit$par["ma1"], fit$par["sigma"])
ans <- fkf(a0 = sp$a0, P0 = sp$P0, dt = sp$dt, ct = sp$ct, Tt = sp$Tt,
           Zt = sp$Zt, HHt = sp$HHt, GGt = sp$GGt, yt = rbind(a))
```

This allows comparision of the prediction with the realization
```{r pred}
plot(ans, at.idx = 1, att.idx = NA, CI = NA)
lines(a, lty = "dotted")
```
and comparision of the filtered series with the realization
```{r filt}
plot(ans, at.idx = NA, att.idx = 1, CI = NA)
lines(a, lty = "dotted")
```

The plotting function for the fkf class also allows visual checking of whether
the residuals are Gaussian
```{r guassian}
plot(ans, type = "resid.qq")
```
or if there is a linear serial dependence through 'acf'
```{r acf}
plot(ans, type = "acf")
```

The smoothed estimates can also be computed by a simple call to fks
```{r fks}
sm <- fks(ans)
```

The smoothed estiamte with confidence intervals can be plotted alongside the
observed data
```{r fks_plot}
plot(sm)
lines(a,col="black")
```
